---
title: "Introduction to probability"
author: "Aarushi Venkatakrishnan"
date: "February 22 2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "#>", highlight = TRUE,
                      fig.align = "center")
```

## Upcoming Deadline

- Exam due by 11:59 PM on February 22nd.
- Lab in groups on Thursday.

## Main ideas

- Use formulas to compute probabilities from tabular data
- Compute empirical probabilities in R via simulation

## Packages

```{r packages}
library(tidyverse)
library(vcd) # used for Arthritis data
```

## Computing probabilities

```{r data}
data(Arthritis)
glimpse(Arthritis)
```

Take a look at the help for `Arthritis` to understand where this data comes
from and the variable meanings.

Let's look at the data in a tabular view. Don't worry about understanding these
functions, we're only using it to better visualize our data via a table.

```{r tabular_data}
xtabs(~ Treatment + Improved, data = Arthritis) %>% 
  addmargins()
```

- How many patients were enrolled in the clinical trial?

84 patients were enrolled in the clinical trial. 

- What is the probability a randomly selected patient received the placebo?

  51.2% (43/84)

- What is the probability a randomly selected patient received the placebo
  and had a marked improvement?
  
  8.3% (7/84)
  
- What is the probability a randomly selected patient received the placebo
  and the treatment?
  
  0% (disjoint probability)
  
- What is the probability a randomly selected patient had some improvement or
  was on the treatment?
  
  57.1% ((7+41)/84)

## Using computer simulations to calculate probabilities

#### Example

Recall that a **vector** is the basic building block in R. Let's create
a vector called `marbles`.

```{r marbles}
marbles <- c("red", "red", "white", "red", "blue", "blue", "red", "blue")
```

Suppose we draw a single marble from our imaginary box, where all the marbles
are equally likely to be selected. What is the probability the marble is blue? 
How about white?

We can simulate this "drawing" with the `sample()` function.

```{r sample}
sample(marbles, size = 1)
```

We produced one random outcome from this experiment. To estimate the probability 
of say getting a white marble, we need to repeat this experiment many many
times.

In the `sample()` function we can change the `size` argument and set
`replace = TRUE`. Setting `replace = TRUE` allows to draw from our population
of eight marbles each time. This way we can easily simulate our marble-drawing 
experiment.

```{r simulate_draws}
draw_results <- sample(marbles, size = 10000, replace = TRUE)
```

```{r summarize_draws}
counts <- table(draw_results)
prop.table(counts)
```

How close is this value to the "true" probability?

This value is very close to the true probability. The true probability of blue 
is 0.375, red is 0.5, and white is 0.125. As we increase the size of our trials,
the law of large numbers indicates that we will get closer to the true
probability.

**To summarize our process:**

1. We defined the sample space for our experiment - `marbles`
2. We simulated this experiment many many times and recorded the outcomes from
   each of the simulations.
3. We computed the relative frequency of the observed outcomes from our many
   simulations.
   
#### Another example

What if we want to compute the probability of getting two marbles of the same
color if we make two draws with replacement? We haven't discussed how to
compute this theoretically yet, but this is what computers are good at.

Before we do this, what is your guess as to what the probability will be?

We'll still use `sample()` to run our simulation many times, but we'll
use `dplyr` functions to compute the relative frequencies.

```{r simulate_two_draws}
two_draw_results <- tibble(
  draw_1 = sample(marbles, size = 10000, replace = TRUE),
  draw_2 = sample(marbles, size = 10000, replace = TRUE)
)
two_draw_results
```

How can we add a variable to `two_draw_results` to see if `draw_1` and `draw_2`
match?

```{r identify_matches}
two_draw_results <- two_draw_results %>% 
  mutate(color_match = draw_1 == draw_2)
two_draw_results
```

All that remains is to compute the relative frequency of the observed outcomes 
from our many simulations.

```{r summarize_matches}
two_draw_results %>% 
  count(color_match) %>% 
  mutate(proportion = n / sum(n))
```

## Practice

Suppose you roll two fair six-sided dice. Which has a higher probability:
the square of dice roll 1 is equal to dice roll 2; or the absolute value of the
difference between dice roll 1 and dice roll 2 is equal to 4.

Perform a simulation to compute this empirical probability.

Write down your guess to the answer before you calculate it.

My prediction is that there is a higher probability fo the second event
occurring because there are more ways that event 2 can be true (dice results 
being 1/5, 2/6, 5/1, 6/2) than event 1 being true (1/1, 2/4).

```{r simulate_rolls}

two_roll_results <- tibble(
  roll_1 = replicate(n = 10000, expr = sample(1:6, size=1)),
  roll_2 = replicate(n = 10000, expr = sample(1:6, size=1))
)
two_roll_results


```

```{r identify_events}

two_roll_results <- two_roll_results %>% 
  mutate(number_match = roll_1^2 == roll_2,
         diff_match = abs(roll_1-roll_2) == 4)
two_roll_results

```

```{r summarize_rolls}
two_roll_results %>% 
  summarize(number_match_prob= mean(number_match), 
            diff_match_prob = mean (diff_match))

```

My prediction was correct. The probability of event 2 happening is almost double
that of the probability of event 1 happening (0.106 vs 0.0574). 


## Additional Resources-please look at before Weds.

- [Open Intro Stats](https://www.openintro.org/book/os/) Sections 3.1 and 3.2
